\section*{Problems}
\addcontentsline{toc}{section}{Problems}


 \begin{prob} \label{prob08.1} Justify your answers to the following: (The setting is a general vector space $V$.)\medskip
\begin{enumerate}[a)] 

\item  Suppose $\uu \in \spn\{\vv,\ww\}$. Show carefully that $\set{\uu,\vv,\ww}$ is linearly dependent.
\medskip
%
\item\sov  Suppose $\uu \in \spn\{\vv,\ww\}$. Show carefully that $\spn\{\vv,\ww\}=\spn\{\uu,\vv,\ww\}$.
\medskip
%

\item  Suppose $\spn\{\vv,\ww\}=\spn\{\uu,\vv,\ww\}$. Show carefully that $\uu \in \spn\{\vv,\ww\}$. 
\medskip
%

\item\sov  Suppose $\spn\{\vv,\ww\}=\spn\{\uu,\vv,\ww\}$. Show carefully that $\set{\uu,\vv,\ww}$ is linearly dependent. 
\medskip
%

\item  Suppose $\spn\{\uu,\vv\}=\spn\{\ww,\xx\}$. Show carefully that $\set{\uu,\vv,\ww}$ and $\set{\uu,\vv,\xx}$ are both linearly dependent. 
\medskip
%
 
\item\sov  Suppose $\set{\vv,\ww}$ is linearly independent, and that $\uu \notin \spn\{\vv,\ww\}$. Show carefully that $\set{\uu,\vv,\ww}$ is linearly independent.
\medskip
% 
\item  Suppose $\set{\vv,\ww}$ is linearly independent, and that $\uu \notin \spn\{\vv,\ww\}$. Show carefully that $\spn\{\vv,\ww\}\not=\spn\{\uu,\vv,\ww\}$.
\medskip
%

\item\sov  Suppose $\spn\{\vv,\ww\}\not=\spn\{\uu,\vv,\ww\}$. Show carefully that $\uu \notin \spn\{\vv,\ww\}$. 
\medskip
% 
\end{enumerate}

\end{prob} \begin{prob} \label{prob08.2} Justify your answers to the following:\medskip
\begin{enumerate}[a)] 


\item Suppose that $\uu,\vv,\ww$ are non-zero vectors in $\R^4$ such that
$\uu\cdot \vv= \uu\cdot \ww= \vv\cdot \ww=0$. Prove that $\set{\uu,\vv,\ww}$ is linearly independent.\medskip



\item\sov Suppose two  polynomials $\set{p, q}$ satisfy $p\not=0$ and $\deg(p) <\deg(q)$. Show carefully that $\set{p,q}$  is linearly independent. \medskip
%



\item Suppose a set of   polynomials $\set{p_1, \dots, p_k}$ satisfies $0\not=p_1$ and $\deg(p_1) <\deg(p_2)< \cdots < \deg (p_k)$. Show carefully that $\set{p_1, \dots, p_k}$  is linearly independent. \medskip
%
\item$^\ast$ Suppose $f,g \in \F(\R)$ are differentiable functions and that  $fg'-f'g $ is not the zero function. Prove carefully that $\set{f,g}$ is linearly independent.\footnote{ Hint: Proceed by contradiction.}\medskip     
%
\item Use the previous part of this question to give another proof (i.e., other the one seen in class) that $\set{\sin x, \cos x}$ is linearly independent.\medskip
%
\item\sov$^\ast$   Suppose $\set{\uu,\vv,\ww}$ are three vectors in $\R^3$ such that $\uu\cdot \vv\times \ww \neq 0$. Prove carefully that $\set{\uu,\vv,\ww}$ is linearly independent.\footnote{  {\it A geometric argument involving `volume' is not sufficient.} Hint: Instead, first recall that  $\set{\uu,\vv,\ww}$ is linearly independent iff none of the vectors is a linear combination of the others. Now proceed by contradiction, and reduce the number of cases to check from 3 to 1 by recalling  from high school that for any three vectors $\vv_1, \vv_2, \vv_3 \in \R^3$ we know that  $\vv_1\cdot \vv_2\times \vv_3=\vv_3\cdot \vv_1\times \vv_2=\vv_2\cdot \vv_3\times \vv_1$.} 
\end{enumerate}
 \end{prob} 
