\section*{Problems}
\addcontentsline{toc}{section}{Problems}




\begin{prob} \label{prob13.1} 

 

\begin{enumerate}[a)]

\item Find the rank of each matrix in Problem \ref{prob12.1}.
\medskip
%
\item\sov Find the ranks of the coefficient matrices and the ranks of the augmented matrices corresponding to the linear systems in Problem \ref{prob12.2}.

\medskip 
%

\end{enumerate}
\end{prob} \begin{prob} \label{prob13.2} Suppose $a,c\in \R$ and consider the following linear system
in the variables
$x,y$ and $z$: 
$$\begin{matrix}  x&+&y &+&az&=&2\\
2x&+&y&+&2a z  &=&3\\ 
3x&+&y &+&3az&=&c \end{matrix}. $$

Note that the general solution of this system may depend on the
values of $a$ and $c$.
Let $[\,A\,|\, b\,]$ denote  the
augmented matrix of the
system above.

\begin{enumerate}[a)]
\item  For 
\underbar{all} values of
$a$ and $c$, find\smallskip
\begin{enumerate}[(i)]
\item 
$\rank A$
\smallskip
\item  $\rank[\,A\,|\, b\,]$  
\end{enumerate}

\item\sov  Find \underbar{all} values of
$a$ and $c$ so that
this system has \smallskip
\begin{enumerate}[(i)]
 
\item a unique solution,\smallskip
\item  infinitely many
solutions,  or \smallskip
\item  
no solutions. 
\end{enumerate}

\medskip
\item In case b (ii) above,
give the general solution, and its complete geometric description. 
\end{enumerate}
\end{prob} \begin{prob} \label{prob13.7} Consider the network of streets with
intersections A, B, C and D below.   The arrows
indicate the direction of traffic flow along the
one-way streets, and the numbers refer to the exact
number of cars observed to enter or leave  A  B,C
and D during one minute.  Each $x_i$
denotes the unknown number of cars which passed
along the indicated streets during the same period.
 $$\xymatrix@=5pt{&&&&&&&&\\&&&&&&&&\\&&&&B\ar[uu]_{10}\ar[rrdd]^{x_4}&&&&\\&&&&&&&&\\
{}\ar[rr]^{11}&&A\ar[rrrr]^{x_5}\ar[rrdd]_{x_2}\ar[rruu]^{x_1}&&&&C\ar[rr]^4
&&\\&&&&&&&&\\&&&&D\ar[rruu]_{x_3}&&&&\\&&&&&&&&\\&&&&\ar[uu]^3&&&&\\}
$$
\begin{enumerate}[a)]
\item  Write down the linear system which describes the the traffic flow, {\bf
together with all the constraints} on the variables $x_i$, $i=1,\dots,5$. (Do not
perform any operations on your equations: this is done for you in (b).)\smallskip

\item The reduced row-echelon form of the augmented matrix from part (a) is 
$$ \bmatrix 
1& 0& 0& -1& 0&|& 10\\  
0& 1& 0& 1&1&|& 1\\ 
0& 0& 1& 1& 1&|& 4\\ 
0& 0& 0& 0& 0&|& 0\endbmatrix$$ Give the general solution. (Ignore the constraints at
this point.)\smallskip

\item  If  AC were closed due to roadwork, find all possible traffic flows, 
 \underbar{using your results from (b).} 

\end{enumerate}

\end{prob} \begin{prob} \label{prob13.3} Consider the network of streets with
intersections A, B, C, D and E below.   The arrows
indicate the direction of traffic flow along the
{\bf one-way streets}, and the numbers refer to the
{\bf exact} number  of cars observed to enter or
leave  A, B, C, D and E during one minute.  Each
$x_i$ denotes the unknown number of cars which
passed along the indicated streets during the 
same period.
$$\xymatrix@=5pt{
&&&&&&&&&&&&\\ 
&&&&&&&&&&&&\\
&&&&&&&&&&&&\\
&&&&&&&&B\ar[rrdd]^{x_4}\ar[uuu]_{50}&&&&\\ 
&&&&&&&&&&&&\\
&&&&&&A\ar[lll]^{60}\ar[rruu]^{x_5}&&&&C\ar[lddd]_{x_3}\ar[llll]_{x_6}&&&&\ar[llll]^{70}\\
&&&&&&&&&&&&\\
&&&&&&&&&&&&\\
&&&&&&&E\ar[luuu]^{x_1}&&D\ar[ll]_{x_2}\ar[dddr]^{60}&&\\  
&&&&&&&&&&&&\\ 
&&&&&&&&&&&&\\
&&&&&&\ar[uuur]^{100}&&&&&&&&\\} 
$$
\smallskip     
\begin{enumerate}[a)]
\item Write down a system of linear equations which describes the the traffic
flow, {  together with all the constraints} on the variables $x_i$,
$i=1,\dots,6$. 

\smallskip

    
\item\sov  The reduced row-echelon form of the augmented matrix of the system in
part (a) is 
$$ \bmatrix 
  1 & 0 & 0 & 0 & -1 & 1 &|& 60 \\
 0 & 1 & 0 & 0 & -1 & 1 &|& -40 \\
 0 & 0 & 1 & 0 & -1 & 1 &|& 20 \\
 0 & 0 & 0 & 1 & -1 & 0 &|&  -50 \\
 0 & 0 & 0 & 0 & 0 & 0 &|& 0
\endbmatrix$$ Give the general
solution. (Ignore the constraints from (a) at this
point.)
\smallskip
 

\item If   $ \overline{ED}$ were closed due to roadwork, find
the minimum flow along  $\overline{AC}$, {\bf using your results from (b).} 
 %
\end{enumerate}

\end{prob} \begin{prob} \label{prob13.4}   State whether each of the following is (always) true,
or is (possibly) false.    
   \smallskip    
\begin{enumerate}[$\bullet$]
\item If you say the statement may be false, you   must give an explicit example.  
\item If you say the statement is true, you must give a clear explanation -   by quoting a theorem presented in class, or by giving a {\it proof valid for every  case}. 
\end{enumerate}
\medskip
\begin{enumerate}[a)]
\item Every non-homogeneous system of 3 equations in 2 unknowns is inconsistent.
\medskip
%
\item\sov Every non-homogeneous system of 3 equations in 2 unknowns is consistent.

\medskip
%
\item Every non-homogeneous system of 2 equations in 3 unknowns has infinitely many solutions.
\medskip
%
\item\sov Every  system of 2 equations in 2 unknowns has a unique solution.
\medskip
%
\item If there is a row of zeroes in the augmented matrix of a linear system, the system will have infinitely many solutions.
\medskip
%
\item\sov If there is a column of zeroes in the coefficient matrix of a consistent linear system, the system will have infinitely many solutions.
\medskip
%
\item If a consistent linear system has infinitely many solutions, there must be a row of zeroes in the reduced augmented matrix (i.e. when the augmented matrix  has been reduced so that the coefficient matrix is in RREF.)
\medskip
%
\item\sov If a consistent linear system has infinitely many solutions, there must be a column of zeroes in the reduced coefficient matrix (i.e. when the augmented matrix has been reduced so that the coefficient matrix is in RREF.)
\medskip
%
\item If the augmented matrix of a homogeneous linear system of 3 equations in 5 unknowns has rank 2, there are 4 parameters in the general solution.
\medskip
%
\item\sov If a homogeneous linear system has a unique solution, it must have the same number of equations as unknowns.
\medskip
%
\end{enumerate}

\end{prob}

