
\begin{sol}{prob13.1} 

 

(b) Find the ranks of the coefficient matrices and the ranks of the augmented matrices corresponding to the linear systems in question \ref{prob12.2}.

\soln



 [\ref{prob12.2}(b)] The rank of the coefficient matrix is 2 and the rank of the augmented matrix is 3. (That's why the corresponding  linear system is inconsistent.)
\medskip

[\ref{prob12.2}(d)] Both ranks are 3.

\medskip 
%
\end{sol}

\begin{sol}{prob13.2} Suppose $a,c\in \R$   and consider the following linear system
in the variables
$x,y$ and $z$: 
$$\begin{matrix}  x&+&y &+&az&=&2\\
2x&+&y&+&2a z  &=&3\\ 
3x&+&y &+&3az&=&c \end{matrix}. $$

Note that the general solution of this system may depend on the 
values of $a$ and $c$. Let $[\,A\,|\, b\,]$ denote  the
augmented matrix of the
system above.

\medskip 
(b)  Find \underbar{all} values of
$a$ and $c$ so that
this system has \smallskip
\begin{enumerate}[(i)]
  
\item a unique solution,\smallskip
\item  infinitely many
solutions,  or \smallskip
\item  
no solutions. 
\end{enumerate}

\soln  

$$[\,A\,|\, b\,] =\bmatrix 1&1&a&|&2\\ 2&1&2a&|&3\\ 3&1&3a&|&c\endbmatrix \sim
\bmatrix 1&1&a&|&2\\ 0&1&0&|&1\\ 0&0&0&|&c-4\endbmatrix $$

Since $\rank A=2<3=\#$ variables,  for {\it all} values of $a,c$, this system {\it never} has a unique solution.
\smallskip

Since $\rank A=2=\rank[\,A\,|\, b\,]$ iff $c=4$, this system will be consistent and have infinitely many
solutions, since then there will be  ($\#\text{variables}-\rank A=3-2=)1$  parameter in the general solution.
\smallskip

Since $\rank A=2<3=\rank[\,A\,|\, b\,]$ iff $c\not=4$, this system will be inconsistent for all values of $c$ except for $c=4$.
\medskip



\end{sol}



\begin{sol}{prob13.3} Consider the network of streets with
intersections A, B, C, D and E below.   The arrows
indicate the direction of traffic flow along the
{\bf one-way streets}, and the numbers refer to the
{\bf exact} number  of cars observed to enter or
leave  A, B, C, D and E during one minute.  Each
$x_i$ denotes the unknown number of cars which
passed along the indicated streets during the 
same period.
$$\xymatrix@=3pt{
&&&&&&&&&&&&\\ 
&&&&&&&&&&&&\\
&&&&&&&&&&&&\\
&&&&&&&&B\ar[rrdd]^{x_4}\ar[uuu]_{50}&&&&\\ 
&&&&&&&&&&&&\\
&&&&&&A\ar[lll]^{60}\ar[rruu]^{x_5}&&&&C\ar[lddd]_{x_3}\ar[llll]_{x_6}&&&&\ar[llll]^{70}\\
&&&&&&&&&&&&\\
&&&&&&&&&&&&\\
&&&&&&&E\ar[luuu]^{x_1}&&D\ar[ll]_{x_2}\ar[dddr]^{60}&&\\  
&&&&&&&&&&&&\\ 
&&&&&&&&&&&&\\
&&&&&&\ar[uuur]^{100}&&&&&&&&\\} 
$$

(b)  The reduced row-echelon form of the augmented matrix of the system in
part (a) is 
$$ \bmatrix 
  1 & 0 & 0 & 0 & -1 & 1 &|& 60 \\
 0 & 1 & 0 & 0 & -1 & 1 &|& -40 \\
 0 & 0 & 1 & 0 & -1 & 1 &|& 20 \\
 0 & 0 & 0 & 1 & -1 & 0 &|&  -50 \\
 0 & 0 & 0 & 0 & 0 & 0 &|& 0
\endbmatrix$$ Give the general
solution. (Ignore the constraints from (a) at this
point.)
\smallskip


\soln  The general solution is 
 
 
\begin{align*} 
x_1&=60 +s-t&\\
x_2&=-40 +s-t&\\
x_3&=20 +s-t& s,t \in \R\\
x_4&=-50 +s&\\
x_5&= s &\\
x_6&=t &
\end{align*}
 

Or it can be written as $$\set{(60 +s-t,-40 +s-t,20 +s-t,-50 +s,s,t)\st st,t \in \R}.$$

 


\end{sol}

\begin{sol}{prob13.4}   State whether each of the following is (always) true,
or is (possibly) false.    
   \medskip    
\begin{enumerate}[$\bullet$]
\item If you say the statement may be false, you   must give an explicit example.  
\item If you say the statement is true, you must give a clear explanation -   by quoting a theorem presented in class, or by giving a {\it proof valid for every  case}. 
\end{enumerate}
\medskip

(b)  Every non-homogenous system of 3 equations in 2 unknowns is consistent.


\soln This is false. For example, the system whose augmented matrix is 
$\bmatrix  
1 & 0 &|& 1 \\
0 & 1 &|& 0\\
0 & 0 &|& 1 \endbmatrix$ is inconsistent.
\medskip
%

(d) Every  system of 2 equations in 2 unknowns has a unique solution.

\soln This is false. For example, the system whose augmented matrix is 
$\bmatrix  
1 & 0 &|& 0 \\
0 & 0 &|& 0\\
 \endbmatrix$ has infinitely many solutions, namely $ (0,s) $ for every $s\in \R$
\medskip

(f) If there is a column of zeroes in the coefficient matrix of a consistent linear system, the system will have infinitely many solutions.

\soln This is true, since once the augmented matrix has been reduced, there will still be a column of zeroes in the coefficient matrix, which means that (as it is assumed consistent), there will be at least one parameter in the general solution.
\medskip

(h) If a consistent linear system has infinitely many solutions, there must be a column of zeroes in the reduced coefficient matrix (i.e. when the augmented matrix has been reduced so that the coefficient matrix is in RRE form.)

\soln This is false. Look at Q.4 (d).
\medskip
%

(j) If a homogenous linear system has a unique solution, it must the same number of equations as unknowns. 

\soln This is false. For example, the system whose augmented matrix is 
$\bmatrix  
1 & 0 &|& 0 \\
0 & 1 &|& 0\\
0 & 0 &|& 0 \endbmatrix$ has a unique solution, namely $(0,0)$.
\medskip
%

\end{sol}

