\section*{Problems}
\addcontentsline{toc}{section}{Problems}

\medskip {\bf Remarks:} 
\begin{enumerate}
\item A question with an asterisk `$ ^\ast$' (or two) indicates a bonus-level question.
 \item You must justify all your responses.  
\end{enumerate}
\bigskip

\centerline{\bf  {Matrix Multiplication}} 
 \begin{prob} \label{prob14.1}
\begin{enumerate}[a)]
\item Find the matrix product $\bmatrix 1&
2\\ 3&6\endbmatrix \bmatrix a&b\\ c&d\endbmatrix$.
\medskip
%
\item\sov Write the matrix product $\bmatrix 1&
2\\ 3&6\endbmatrix \bmatrix a\\ b\endbmatrix$ as a linear combination of the columns of $A=\bmatrix 1&
2\\ 3&6\endbmatrix$.
\medskip
%
\item Write the matrix product $\bmatrix 1&
2\endbmatrix \bmatrix a&b\\ c&d\endbmatrix$ as a linear combination of the rows of $B=\bmatrix a&b\\ c&d\endbmatrix$.
\medskip
%
\item\sov  Find the matrix product $\bmatrix a \\ b \endbmatrix \bmatrix c&d \endbmatrix$.
\medskip
%

\item Find the matrix product $\bmatrix c&d \endbmatrix \bmatrix a \\ b \endbmatrix $.
\medskip
%
\item\sov  If $A=\bmatrix c_1&c_2 &c_3\endbmatrix$ is an $m\times 3$ matrix written in block column form, and $x=\bmatrix 2\\1\\ 4\endbmatrix$ is a column vector in $\R^3$, express $Ax$ as a linear combination of $c_1, c_2$ and $c_3$. 
\medskip
%
\item Show that $
\bmatrix 1&2\\ 3&6\endbmatrix \bmatrix 2&-4\\ -1&2\endbmatrix=\bmatrix0&0\cr 0&0\endbmatrix$.
\medskip
%
\item\sov  Show that if $A=
\bmatrix 0&1\\ 0&0\endbmatrix$, then $A^2=\bmatrix0&0\cr 0&0\endbmatrix$. 
\medskip
%
\item Show that if $A=
\bmatrix 0&1\\ 0&0\endbmatrix$, and $B=\bmatrix 0&0\\ 1&0\endbmatrix$ then $AB \not= BA$. 
\medskip
%
\item\sov  If $C$ is a $m\times 4$ matrix and $D  =\bmatrix 1&0&1\\0&1&1\\ 1&0&0\\0&0&0 \endbmatrix $, express the columns of $CD$ in terms of the columns of $C$.
\medskip
%
\item If $A=\bmatrix 1&0&1\\0&1&1\\ 1&0&0\\0&0&0 \endbmatrix  $ and $B$ is a $3\times n$ matrix   , express the rows of $AB$ in terms of the rows of $B$
\medskip
%
\item\sov   Find  all $ (a,\ b,\ c)$ so that $\bmatrix 1&
2\\ 3&6\endbmatrix \bmatrix a&b\\ c&a\endbmatrix=\bmatrix0&0\cr 0&0\endbmatrix
 $.
\medskip
%
\item Compute $ \bmatrix  1&0&0&-3\\ 0&1&3&0\\ 0&0&1&0
\\ 0&0&0&1
\endbmatrix^{2015}$
\medskip
%
\end{enumerate}
 \end{prob}
\begin{prob}
\label{prob14.2}
Show the formula $A^2 - \tr(A)A + \det(A)I = 0$ holds for any $2\times 2$ matrix $A$.
\end{prob}

\begin{prob} \label{prob14.3} State whether each of the following is (always) true,
or is (possibly) false.     In this question, $A$ and $B$ are matrices for which the indicated products exist.
   \smallskip    
\begin{enumerate}[$\bullet$]
\item If you say the statement may be false, you must give an explicit example.   
\item If you say the statement is true, you must give a clear explanation -   by quoting a theorem presented in class, or by giving a {\it proof valid for every  case}. 
\end{enumerate}
\medskip
\begin{enumerate}[a)]
\item  $(A+B)^2= A^2 + 2 AB + B^2$
\medskip
%
\item\sov $C(A+B)=CA+CB$
\medskip
%
\item $(A+B)I_6=A+B$
\medskip
%
\item\sov $AB=BA$
\medskip
%
\item $(AB)C=A(BC)$
\medskip
%
\item\sov  If $A^2=0$ for a square matrix $A$, then $A=0$.
\medskip
%
\end{enumerate}
 
\end{prob} \begin{prob} \label{prob14.4}$^\ast$ Suppose $A$ is a {\it symmetric} $n \times n$ matrix, i.e., $A=A^t$. Show that if $v$ and $w$ are any vectors in $\R^n$, then $(Av)\cdot w= v \cdot (Aw)$.\end{prob}\footnote{ Hint: Remember that if we write vectors as columns, then the dot product $x\cdot y$ is the same as the matrix product $x^t y$. Write the dot products as matrix products and expand.}

\bigskip

\centerline{\bf  {Applications to Linear Systems}} 
 \begin{prob} \label{prob14.5} Write the matrix equation which is equivalent to each of the following linear systems.
\medskip
\begin{enumerate}[a)]
\item 
$$\begin{matrix} 
x&+&y&+&z&=&0\\
-9x&-&2y&+&5z&=&0\\
-x&+&y&+&3z&=&0\\
-7x&-&2y&+&3z&=&0 \end{matrix} $$
\medskip
%

\item\sov $$\begin{matrix} x&&&&&+&w&=&1\\
x&&&+&z&+&w&=&0\\
x&+&y&+&z&&&=&-3\\
x&+&y&&&-&2w&=&2 \end{matrix} $$
\medskip
% 
\item 
$$\begin{matrix} &&&-&2x_{3}&&&+&7x_{5}&=12\\
2x_{1}&+&4x_{2}&-&10x_{3}&+&6x_{4}&+&12x_{5}&=28\\
2x_{1}&+&4x_{2}&-&5x_{3}&+&6x_{4}&-&5x_{5}&=-1\end{matrix}$$
\medskip
%
\end{enumerate}
 

\end{prob} \begin{prob} \label{prob14.6}  Write the matrix equation of the linear system corresponding to each of the augmented matrices given below.
\medskip
\begin{enumerate}[a)]

\item $\bmatrix  0 & 1 &|& 0 \\
 0 & 0 &|& 0 \endbmatrix$
\medskip
%
\item\sov $ \bmatrix 1 & 0 & -1 &|&0 \\
 0 & 1 & 2 &|&0\\
 0 & 0 & 0&|&1 \\
 0 & 0 & 0 &|&0\endbmatrix$
\medskip
%
\item $\bmatrix  1 & 0 & 0 & 3 &|& 3 \\
 0 & 1 & 0 & 2 &|& -5 \\
 0 & 0 & 1 & 1 &|& -1 \\
 0 & 0 & 0 & 0 &|& 0\endbmatrix$
\medskip
%
\item\sov $\bmatrix  1 & 2 & 0 & 3 & 0 &|& 7 \\
 0 & 0 & 1 & 0 & 0 &|& 1 \\
 0 & 0 & 0 & 0 & 1 &|& 2 \endbmatrix$
\medskip
%

\item $ \bmatrix 
1& 0& 0& -1& 0&|& 10\\  
0& 1& 0& 1&1&|& 1\\ 
0& 0& 1& 1& 1&|& 4\\ 
0& 0& 0& 0& 0&|& 0\endbmatrix$
\medskip
%
\end{enumerate}

\end{prob} 

 \begin{prob} \label{prob14.7} State whether each of the following is (always) true,
or is (possibly) false.     The matrices are assumed to be square.  
   \smallskip    
\begin{enumerate}[$\bullet$]
\item If you say the statement may be false, you must give an explicit example.   
\item If you say the statement is true, you must give a clear explanation -   by quoting a theorem presented in class, or by giving a {\it proof valid for every  case}. 
\end{enumerate}
\medskip
\begin{enumerate}[a)]
\item If $[A\,|\,\bb\,]$ is the augmented matrix of a linear system, then $\rank A > \rank  [A\,|\,\bb\,]$ is possible.
\medskip
%
\item\sov  If $[A\,|\,\bb\,]$ is the augmented matrix of a linear system, then $\rank A<\rank  [A\,|\,\bb\,]$ is possible.
\medskip
%

\item If $[A\,|\,\bb\,]$ is the augmented matrix of a linear system, and $\rank A=\rank  [A\,|\,\bb\,]$, then the system  is inconsistent.\medskip %(inconsistent or consitent???)

%
\item\sov  If $[A\,|\,\bb\,]$ is the augmented matrix of a linear system, and $\rank A=\rank  [A\,|\,\bb\,]$, then the system  is consistent. 
 
%
\medskip
\item If $[A\,|\,\bb\,]$ is the augmented matrix of a linear system, and $\rank A=\rank  [A\,|\,\bb\,]$, then the vector $\bb$ is a linear combination of the columns of $A$. 
\medskip
%
\item\sov  If $A$ is an $m \times n$ matrix and $A\xx=\zero$ has a unique solution for $\xx\in \R^n$, then the columns of $A$ are linearly independent.
%
%\medskip
\item If $A$ is an $m \times n$ matrix and $A\xx=\zero$ has a unique solution for $\xx\in \R^n$, then the rows of $A$ are linearly independent.
%
\item\sov  If $A$ is an $m \times n$ matrix and $A\xx=\zero$ has infinitely many solutions for $\xx\in \R^n$, then the columns of $A$ are linearly dependent.
%
\medskip
\item If $A$ is an $m \times n$ matrix and $A\xx=\zero$ has infinitely many solutions for $\xx\in \R^n$, then the rows of $A$ are linearly dependent.
%
\medskip
\item\sov  If $A$ is a $6 \times 5$ matrix and $\rank A=5$, then $A\xx=\zero$ implies $\xx=\zero$ for $\xx\in \R^5$
\medskip
%
\item If $A$ is a $6 \times 5$ matrix and $\rank A=5$, then $A\xx=\bb$ is consistent for every $\bb \in \R^6$.
\medskip
%
\item\sov  If $A$ is a $5 \times 6$ matrix and $\rank A=5$, then $A\xx=\bb$ is consistent for every $\bb \in \R^5$.
\medskip
%
\item If $A$ is a $5 \times 6$ matrix and $\rank A=5$, then $A\xx=\zero$ implies $\xx=\zero$ for $\xx\in \R^6$.
\medskip
%
\item\sov  If $A$ is a $3 \times 2$ matrix and $\rank A=1$, then $A\xx=\zero$ implies $\xx=\zero$ for $\xx\in \R^2$.
\medskip
%
\item The columns of a $19\times 24$ matrix are
always linearly dependent.
\medskip
%
\item\sov  The rows of a $19 \times 24$ matrix are
always linearly dependent.
\medskip
%
\end{enumerate}



\end{prob}

